---
title: "SpaceANOVA"
author: Souvik Seal
output: 
 github_document: 
  pandoc_args: --webtex=http://chart.apis.google.com/chart?cht=tx&chl=
bibliography: ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  out.width = "100%",
  messages = FALSE,
  warnings = FALSE
)
```

This is an R package implementing the proposed method from the paper, "SpaceANOVA: Spatial co-occurrence analysis of cell types in multiplex imaging data using point
process and functional ANOVA". It can be used to study differential spatial co-occurrence of pairs of cell types across multiple groups, such as case/control.

## Loading the package

We install and load the developmental version of SpaceANOVA from GitHub.

```{r loading packages, message=FALSE}

suppressMessages(devtools::install_github('sealx017/SpaceANOVA'))
require(SpaceANOVA)

```

## Loading the example dataset

Next, we import the type 1 diabetis melitus (T1DM) IMC dataset [1].

```{r loading the marker expression, out.width = "50%"}

data("IMC_T1DM")
knitr::kable(head(IMC_T1DM), format="markdown")

```

## Perform the analysis

* We start by showing how to compute the MI-based measures, EQMI\*, EQMI, and CSQMI (Principe 2010) between any arbitrary set of random variables (r.v.'s) from a general data-frame i.e., when the data does not necessarily come from multiplex imaging platforms. We create a matrix named Data_matrix with 2500 samples and 5 columns corresponding to five r.v.'s and use it in the function named QMI to estimate the measures.



```{r Computing EQMI*, EQMI and CSQMI with an arbitrary data matrix, echo = T, results = T}

Final_result = All_in_one(data = IMC_T1DM, fixed_r = seq(0, 100, by = 1), Summary_function = "g", Hard_ths = 10, perm = TRUE, nPerm = 20,  print = T)

```

## Computing subject specific EQMI* values with the imaging data

We return to the analysis of multiplex imaging data. We compute the EQMI\* of all the subjects using the function QMI_all and store the values in a matrix whose every row corresponds to a subject. 

```{r EQMI_star_computation}

EQMI_vector = QMI_all(marker_data, bandwidth = "Ind", measure = "EQMI_star", progress_bar = "False")
knitr::kable(head(EQMI_vector), format="markdown")

```

## Association testing using CoxPH model

### Association testing with survival outcome

Using the vector of estimated EQMI\* of all the subjects, we perform association analysis with two clinical outcomes, survival and recurrence. We have the outcomes, the time to death and the time to recurrence and the respective censoring indicators (= 0 for an event) as the columns of the matrix named clinical_data. We have one single covariate, Age in the same matrix. Below, we create a matrix named surv_dat with all the subject IDs and their survival outcomes, and another matrix named covariates with the subject IDs and available covariates which, in this case, is just Age. We then use the function Cox_PH to fit the proportional hazard (PH) model outputting a table of p-values. To add higher order terms in the PH model, change degree to > 1. 

```{r CoxPh model studying association between EQMI and survival}

surv_dat = clinical_data[,c(1,4:5)]
covariates = clinical_data[,c(1,6)]
SurvCox = Cox_PH(surv_dat, covariates, EQMI_vector, degree = 1)
knitr::kable(SurvCox, format="markdown")

```

### Association testing with recurrence outcome

Below, we create a matrix named recur_dat with all the subject IDs and their recurrence outcomes, and use the same covariates matrix as earlier. We again use the Cox_PH function and get the p-values.

```{r CoxPh model studying association between EQMI and recurrence}

recur_dat = clinical_data[,c(1:3)]
RecurCox = Cox_PH(recur_dat, covariates, EQMI_vector, degree = 1)
knitr::kable(RecurCox, format="markdown")

```

## Additional tools: Compute and plot univariate and bivariate kernel densities of the markers

#### Univariate marginal density

We provide a few basic functions to estimate and plot the univariate marginal densities of the random variables. We go back to the matrix named Data_matrix which has 2500 samples and 5 columns corresponding to five r.v.'s. We look at first two columns (call them, X_1 and X_2) and estimate their kernel density estimates as f_1 and f_2 using the function dens_univ. Next, using the function univ_dens_plot, we plot f_1 and f_2. The default number of grids is, ngrids = 1024 and the bandwidth parameter is selected using Silverman's rule.

```{r Computing univ density, fig.show = "hold", out.width = "50%"}

X_1 = Data_matrix[,1]
X_2 = Data_matrix[,2]  

f_1 = univ_dens(X_1, ngrids = 1024)
f_2 = univ_dens(X_2, ngrids = 1024)

p1 = univ_dens_plot(f_1)
p2 = univ_dens_plot(f_2)

```

#### Bivariate joint density

We estimate the bivariate joint density of X_1 and X_2, as f_12 using the function biv_dens. Next, using the function biv_dens_plot, we plot f_12. The default number of grids is, ngrids = 512 and the bandwidth matrix used is the multivariate plug-in bandwidth matrix described in Wand, M. P., & Jones, M. C. (1994). It may be easy to interpret the estimated density focusing only on the smaller values of $X_1$ and $X_2$ since both the r.v.'s usually have most of the values close to 0. Hence, we use two thresholds (q1, q2) to display the estimated density only between 70\% quantiles of both the r.v.'s. 

```{r Computing biv density}


f_12 = biv_dens(cbind(X_1, X_2), ngrids = 512)
q1 = quantile(X_1, 0.7); q2 = quantile(X_2, 0.7)
p = biv_dens_plot(f_12, maxs = c(q1, q2))


```


## References

1\) Damond, N., Engler, S., Zanotelli, V. R., Schapiro, D., Wasserfall, C. H.,
Kusmartseva, I., Nick, H. S., Thorel, F., Herrera, P. L., Atkinson, M. A., et al.
(2019). A map of human type 1 diabetes progression by imaging mass cytometry.
Cell metabolism, 29(3), 755–768.

2\) Baddeley, A., Rubak, E., and Turner, R. (2015). Spatial point patterns: methodology and applications with R. CRC press.

3\) Zhang, J.-T. (2013). Analysis of variance for functional data. CRC press.



