---
title: "SpaceANOVA"
author: Souvik Seal
output: 
 github_document: 
  pandoc_args: --webtex=http://chart.apis.google.com/chart?cht=tx&chl=
bibliography: ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  out.width = "100%",
  messages = FALSE,
  warnings = FALSE
)
```

This is an R package implementing the proposed method from the paper, "SpaceANOVA: Spatial co-occurrence analysis of cell types in multiplex imaging data using point
process and functional ANOVA". It can be used to study differential spatial co-occurrence of pairs of cell types across multiple groups, such as case/control.

## Loading the package

We install and load the developmental version of SpaceANOVA from GitHub.

```{r loading packages, message=FALSE}

suppressMessages(devtools::install_github('sealx017/SpaceANOVA'))
require(spatstat)
require(SpaceANOVA)

```

## Loading the example dataset

Next, we import the type 1 diabetis melitus (T1DM) IMC dataset [1].

```{r loading the marker expression, out.width = "50%"}

data("IMC_T1DM")
knitr::kable(head(IMC_T1DM), format="markdown")

```

## Perform the analysis

* We start by showing how to compute the MI-based measures, EQMI\*, EQMI, and CSQMI (Principe 2010) between any arbitrary set of random variables (r.v.'s) from a general data-frame i.e., when the data does not necessarily come from multiplex imaging platforms. We create a matrix named Data_matrix with 2500 samples and 5 columns corresponding to five r.v.'s and use it in the function named QMI to estimate the measures.



```{r Computing EQMI*, EQMI and CSQMI with an arbitrary data matrix, echo = T, results = T}

Final_result = All_in_one(data = IMC_T1DM, fixed_r = seq(0, 100, by = 1), Summary_function = "g", Hard_ths = 10, perm = TRUE, nPerm = 20,  print = T)

```

## Computing subject specific EQMI* values with the imaging data

We return to the analysis of multiplex imaging data. We compute the EQMI\* of all the subjects using the function QMI_all and store the values in a matrix whose every row corresponds to a subject. 

```{r EQMI_star_computation}


```

## Association testing using CoxPH model

### Association testing with survival outcome

Using the vector of estimated EQMI\* of all the subjects, we perform association analysis with two clinical outcomes, survival and recurrence. We have the outcomes, the time to death and the time to recurrence and the respective censoring indicators (= 0 for an event) as the columns of the matrix named clinical_data. We have one single covariate, Age in the same matrix. Below, we create a matrix named surv_dat with all the subject IDs and their survival outcomes, and another matrix named covariates with the subject IDs and available covariates which, in this case, is just Age. We then use the function Cox_PH to fit the proportional hazard (PH) model outputting a table of p-values. To add higher order terms in the PH model, change degree to > 1. 

```{r CoxPh model studying association between EQMI and survival}


```

## References

1\) Damond, N., Engler, S., Zanotelli, V. R., Schapiro, D., Wasserfall, C. H.,
Kusmartseva, I., Nick, H. S., Thorel, F., Herrera, P. L., Atkinson, M. A., et al.
(2019). A map of human type 1 diabetes progression by imaging mass cytometry.
Cell metabolism, 29(3), 755–768.

2\) Baddeley, A., Rubak, E., and Turner, R. (2015). Spatial point patterns: methodology and applications with R. CRC press.

3\) Zhang, J.-T. (2013). Analysis of variance for functional data. CRC press.



