---
title: "SpaceANOVA"
author: Souvik Seal
output: 
 github_document: 
  pandoc_args: --webtex=http://chart.apis.google.com/chart?cht=tx&chl=
bibliography: ref.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  out.width = "100%",
  messages = FALSE,
  warnings = FALSE
)
```

This is an R package implementing the proposed method from the paper, "SpaceANOVA: Spatial co-occurrence analysis of cell types in multiplex imaging data using point
process and functional ANOVA". It can be used to study differential spatial co-occurrence of pairs of cell types across multiple groups, such as case/control.

## Loading the package

We install and load the developmental version of SpaceANOVA from GitHub.

```{r loading packages, message=FALSE}

suppressMessages(devtools::install_github('sealx017/SpaceANOVA'))
require(spatstat)
require(tidyr)
require(dplyr)
require(fda.usc)
require(ggplot2)
require(cowplot)
require(SpaceANOVA)

```

## Loading the example dataset

Next, we import the type 1 diabetis melitus (T1DM) IMC dataset [1]. The input data should have a simple form: a) rows should correspond to individual cells, and b) six columns should correspond to the group, subject, and image indices of the cell, along with it's x,y co-ordinates and cell type. 

```{r loading the marker expression, out.width = "50%"}

data("IMC_T1DM")
knitr::kable(head(IMC_T1DM), format="markdown")

```

## Compute the summary functions and perform association analysis

This function can be used to perform the entire analysis at one go, starting from summary function estimation to the univariate and multivariate FANOVA-based association analyses. 
The function takes several input parameters out of which the most important ones are displayed below. "fixed_r" corresponds to the grid values of radius r. "Summary_function" corresponds to which summary function to be used: g, K, or L.  "Hard_ths" corresponds to the lowest number of a particular cell type, say A, that an image can have (the image will be dropped if not for any pair of cell types involving A). "perm" is used to employ the permutation envelope-based adjustment to the summary functions in presence of holes in images and "nPerm" denotes the number of permutations to be used. "cores" is the number of cores to be used, if more than 1, mclapply (parallel package) is called. 


```{r Computing spatial summary functions, and performing both univariate and multivariate differential co-occurrence analysis, echo = T, results = T}

Final_result = All_in_one(data = IMC_T1DM, fixed_r = seq(0, 100, by = 5), Summary_function = "g", Hard_ths = 10, perm = TRUE, nPerm = 20, cores = 8)

```

## Extracting p-values 

Extract the p-values of the univariate and multivariate tests: SpaceANOVA Univ. and SpaceANOVA Mult. and print.

```{r P-value extraction}
p_res = p_extract(Final_result)
Univ_p = p_res[[1]]
Mult_p = p_res[[2]]
print(Univ_p)

```

## Heatmap of -log10 of p-values 


Display the -log10 of the p-values using heatmap.

```{r P-value visualization}

Plot.heatmap(Univ_p, main = "SpaceANOVA Univ.")
Plot.heatmap(Mult_p, main = "SpaceANOVA Mult.")

```

## Visualizing summary functions

Next, we can check the summary functions corresponding to those pairs of cell types whose p-values turned out to be significant. In this case, we visualize the g-functions of pairs: (alpha, alpha) and (beta, Th). We also display the point-wsie F-values which might help to understand which values of the radius r were most influential.

```{r plotting summary functions}

Pointwise_Fvals = Final_result[[1]][[2]]
Functional_results = Final_result[[2]]

Plot.functions(Functional_results, pairs = c("alpha", "alpha"), Fvals = Pointwise_Fvals)

```

## Visualizing cellular organization

We can also visualize cellular organization in different images of every subject. 

```{r plotting celluar organization}

```


## References

1\) Damond, N., Engler, S., Zanotelli, V. R., Schapiro, D., Wasserfall, C. H.,
Kusmartseva, I., Nick, H. S., Thorel, F., Herrera, P. L., Atkinson, M. A., et al.
(2019). A map of human type 1 diabetes progression by imaging mass cytometry.
Cell metabolism, 29(3), 755–768.

2\) Baddeley, A., Rubak, E., and Turner, R. (2015). Spatial point patterns: methodology and applications with R. CRC press.

3\) Zhang, J.-T. (2013). Analysis of variance for functional data. CRC press.



